---
title: "STAT/MATH 495: Problem Set 06"
author: "Albert Y. Kim and Andrew Kim"
date: "2017-10-17"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    collapsed: false
    smooth_scroll: false
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, fig.width=8, fig.height=4.5, message=FALSE, warning = FALSE
  )
set.seed(76)

# Load packages
library(tidyverse)
library(broom)
library(knitr)
```





# Collaboration

Please indicate who you collaborated with on this assignment: 





# Setup

Define truth, which again we know for the purposes of this assignment, but in
practice we won't:

* the true function f(x) i.e. the signal
* the true epsilon i.e. the noise, which in this case is Normal$(0, sd=\sigma)$.
Hence the standard deviation $\sigma$ determines the amount of noise.

```{r}
f <- function(x) {
  x^2
}
sigma <- 0.3
```

This is the target point we'll be trying to predict: $(0.95, f(0.95)) = (0.95, 0.95^2) = (0.95, 0.9025)$, Thus, the test set is just `x=0.95`

```{r}
x0 <- 0.95
test_set <- data_frame(x=x0)
```

This function generates a random sample of size $n$; think of this as a "get new
data" function. Random in terms of both:

* (New) the predictor x (uniform on [0,1])
* the amount of noise $\epsilon$

```{r}
generate_sample <- function(f, n, sigma) {
  sample <- data_frame(
    x = runif(n = n, min = 0, max = 1),
    f_x = f(x),
    epsilon = rnorm(n = n, mean = 0, sd = sigma),
    y = f_x + epsilon
  )
  # Recall: We don't observe f(x) and epsilon, just (x, y)
  sample <- sample %>% 
    select(x, y)
  
  return(sample)
}
```

Define

* The number $n$ of observations $(x_i, y_i)$ in each sample. In the handout,
$n=100$ to keep plots uncrowded. Here we boost to $n=500$
* Number of samples of size $n$ to consider

```{r}
n <- 500
n_samples <- 10000
```


# Computation

```{r, cache=TRUE}
# Save predictions here
y_hat_lm <- rep(0, n_samples)
y_hat_splines_99 <- rep(0, n_samples)

for(i in 1:n_samples) {
  # 1. Sample training data
  train_set <- generate_sample(f, n, sigma)

  # 2.a) Fit lm model
  lm_model <- lm(y ~ x, data = train_set)
  # 2.b) Make predictions
  y_hat_lm[i] <- lm_model %>%
    augment(newdata=test_set) %>%
    pull(.fitted)

  # 3.a) Fit splines model
  splines_model <- smooth.spline(x=train_set$x, y=train_set$y, df=99) 
  y_hat_splines_99[i] <- predict(splines_model, x=test_set$x) %>% 
    as_tibble() %>% 
    pull(y)
}
```

```{r, cache=TRUE}
# Generate truth: i.e. all components we really know, but pretend not to:
truth <- data_frame(
  x = rep(x0, n_samples),
  f_x = f(x),
  eps = rnorm(n = n_samples, mean = 0, sd = sigma),
  y = f_x + eps
)

# Show bias^2/variance/irreducible error breakdown for lm:
results1 <- truth %>%
  mutate(y_hat = y_hat_lm) %>% 
  summarise(
    MSE = mean((y-y_hat)^2),
    bias_squared = mean((f_x-y_hat))^2,
    var = var(y_hat)
  ) %>%
  mutate(
    irreducible = sigma^2,
    sum = bias_squared + var + irreducible
    ) %>% 
  kable(digits = 4)

# Show bias^2/variance/irreducible error breakdown for splines:
results2 <- truth %>%
  mutate(y_hat = y_hat_splines_99) %>% 
  summarise(
    MSE = mean((y-y_hat)^2),
    bias_squared = mean((f_x-y_hat))^2,
    var = var(y_hat)
  ) %>%
  mutate(
    irreducible = sigma^2,
    sum = bias_squared + var + irreducible
  ) %>% 
  kable(digits = 4)
```

# Tables

As done in Lec 2.7, for both

* An `lm` regression AKA a `smooth.splines(x, y, df=2)` model fit 
* A `smooth.splines(x, y, df=99)` model fit 

output tables comparing:

|  MSE| bias_squared|   var| irreducible|   sum|
|----:|------------:|-----:|-----------:|-----:|
|     X|           X  |     X |      X |         X |

where `sum = bias_squared + var + irreducible`. You can created cleanly formatted tables like the one above by piping a data frame into `knitr::kable(digits=4)`.


### Output for linear regression model results:

```{r, echo = FALSE}
results1
```


### Output for spline model results:

```{r, echo = FALSE}
results2
```


# Analysis

**Questions**:

1. Based on the topics covered in Lec 2.7, name one possible "sanity check" for your results. Name another if you can.
1. In **two** sentences or less, give a rough sketch of what the procedure would
be to get the breakdown of $$\mbox{MSE}\left[\widehat{f}(x)\right]$$ for *all*
$x$ in this example, and not just for $$\mbox{MSE}\left[\widehat{f}(x_0)\right]
= \mbox{MSE}\left[\widehat{f}(0.95)\right]$$.
1. Which of the two models would you choose for predicting the point of interest and why?

**Answers**:

1.
2.
3. Choose the spline model because it's has a better balance of bias and variance compared to linear regression. The linear regression model is much more biased compared to the spline model.
